ChemKED: a human- and machine-readable data standard for chemical kinetics experiments

### 2017-02-26 pyked.tex
draft of intro/related work section

%====================================================================
\section{Introduction}
%
Fundamental combustion experiments provide vital data for understanding
fuel chemistry and validating chemical kinetic models. Important measured quantities
include autoignition delay times, laminar flame speeds, and species profiles.
However, despite both the importance and abundance of such information in the literature,
the community lacks a widely adopted standard format for this data. This impedes
both sharing and wide use by the community.

%====================================================================
\section{Related work}
%
\todo[inline]{Kyle to do}

The combustion\slash chemical kinetics community lacks an accepted, commonly used
standard for recording and sharing data from fundamental combustion experiments.
Interestingly, this contrasts the situation for chemical kinetic models, where the
CHEMKIN~\autocite{Kee:1996ck} format dominates. Competing standards such as the
Cantera~\autocite{Cantera:2.3.0} CTI format or FlameMaster~\autocite{FlameMaster:ref,FlameMaster:code}
lag behind considerably, although use of the former continually grows due to Cantera's
open availability. Similarly,

Instead, such data must typically be found in under-documented, nonstandard comma-separated value
files and Excel spreadsheets, or even contained in PDF tables or figures rather than as
supplementary material associated with a paper.

While the community has not widely adopted a standard format,
---most notably, Frenklach developed the PrIMe (\textbf{Pr}ocess \textbf{I}nformatics
\textbf{M}od\textbf{e}l) data format~\autocite{Frenklach:2007bm,You:2011hy} and an associated
online database, the PrIMe Data Warehouse~\autocite{PrIMe}.

More recently, Varga et al.~\autocite{Varga2015a,Varga2015b} developed the \texttt{ReSpecTh} standard,
an XML-based format that builds on, and is compatible with, PrIMe.

We aim to develop a new, human- and machine-readable data standard for fundamental combustion
experiments, \ck{}, and offer tools for easily working with data encoded in this format.
Our motivations are similar to some of those of the PrIMe and \texttt{ReSpecTh} teams, although
less ambitious in scope: enable easy sharing and use of data, for the (primary) purposes of
developing and validating predictive chemical kinetic models.
However, those efforts were generally less successful in being adopted---we hope to succeed
by developing more user-friendly data files and tools for interacting with them.

Niemeyer recently introduced an initial version of the \ck{} format~\autocite{Niemeyer:2016wf},
and in this work we further formalize and develop the standard. We also discuss a related
Python software tool for validating and working with \ck{} files: \pk{}.

>>>>

%====================================================================
\section{Introduction}
%
Fundamental combustion experiments provide vital data for understanding
fuel chemistry and validating chemical kinetic models. Important measured quantities
include autoignition delay times, laminar flame speeds, and species profiles.
However, despite both the importance and abundance of such information in the literature,
the combustion\slash chemical kinetics community lacks an accepted, commonly used
standard for recording and sharing data from fundamental combustion experiments.

Interestingly, this contrasts the situation for chemical kinetic models, where the
CHEMKIN~\autocite{Kee:1996ck} format dominates. Competing standards such as the
Cantera~\autocite{Cantera:2.3.0} CTI format or FlameMaster~\autocite{FlameMaster:ref,FlameMaster:code}
lag behind considerably, although use of the former continually grows due to Cantera's
open availability.
% Similarly, % with more space, discuss JANAF/NASA polynomial standard ?

Instead, such data must typically be found in non- or under-documented, nonstandard
comma-separated value files and Excel spreadsheets, or contained in PDF tables
rather than as supplementary material associated with a paper.
(In the worst cases, data is only available in figures---and in some cases not at all.)
\todo{call out specific examples of these...?}{}
These practices limit the wide use of the valuable, extensive experimental data
available in the literature. Frenklach~\autocite{Frenklach:2007bm} further explained the
benefits of standardized, widely available data for combustion.

While the community has not widely adopted a standard format or the practice of openly
sharing data, some researchers and groups attempted to solve the first problem.
Most notably, Frenklach developed the PrIMe (\textbf{Pr}ocess \textbf{I}nformatics
\textbf{M}od\textbf{e}l) data format~\autocite{Frenklach:2007bm,You:2011hy} and an associated
online database, the PrIMe Data Warehouse~\autocite{PrIMe}. PrIMe files encode fundamental
combustion experimental data, along with kinetic models and calculated quantities,
using the Extensible Markup Language (XML) standard.

PrIMe represented an admirable first attempt at standardization, but a number of limitations
prevented wide adoption: the PrIMe standard does not require or support all information
needed to simulate an experiment, including a machine-readable definition of ignition or a
standard way to express the rapid compression machine (RCM) compression process; and
the XML format, intended more as a markup language than data format, presents a barrier
to creating and working with files.

More recently, Varga et al.~\autocite{Varga2015a,Varga2015b} developed the ReSpecTh standard,
an XML-based format that builds on PrIMe. While ReSpecTh adds important features, it still
suffers from the same usability issues as PrIMe.

We aim to develop a new, human- and machine-readable data standard for fundamental combustion
experiments, \ck{}, and offer tools for easily working with data encoded in this format.
Niemeyer recently introduced an initial version of the \ck{} format~\autocite{Niemeyer:2016wf},
and in this work we further formalize and develop the standard. We also discuss a related
Python software tool for validating and working with \ck{} files: \pk{}.
Our motivations are similar to some of those of the PrIMe and ReSpecTh teams, although
less ambitious in scope: enable easy sharing and use of data, for the (primary) purposes of
developing and validating predictive chemical kinetic models.
However, those efforts were generally less successful in being adopted---we hope to succeed
by developing more user-friendly data files and tools for interacting with them.
Furthermore, we plan to share all data and software openly, and cultivate a community of
user-developers.
